BayesPy: Variational Bayesian Inference in Python

### 2015-06-01 bayespy.tex
Remove tikz and bayesnet dependency
Minor fixes by reviewers

\usepackage{subcaption}
\usepackage{url}
\usepackage{color}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{bayesnet}

@@

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct, for instance,
principal component analysis models, linear state-space models, mixture models
and hidden Markov models.  In addition, BayesPy has been used in two
publications about parameter expansion and time-varying dynamics for linear
state-space models \citep{Luttinen:2013,Luttinen:2014}.
% , which show that it is applicable to large complex models requiring efficient
% inference.

@@

Nodes are the key part of BayesPy as they are used to construct the model.
BayesPy has two types of nodes: stochastic and deterministic.  Stochastic nodes
correspond to probability distributions and deterministic nodes correspond to
functions.  Built-in stochastic nodes include all common exponential family
distributions (e.g., Gaussian, gamma and Dirichlet distributions), a general
mixture distribution and a few complex nodes for dynamic variables (e.g.,
discrete and Gaussian Markov chains).  Built-in deterministic nodes include a
gating node and a general sum-product node.
%The
%documentation also provides instructions for implementing new nodes in case the
%built-in collection is insufficient.
In case a model cannot be constructed using the built-in nodes, the
documentation provides instructions for implementing new nodes.

BayesPy is designed to be simple enough for average users but flexible and
efficient enough for expert users.  One goal is to keep the syntax easy and
intuitive to read and write by making it similar to the mathematical formulation
of the model.  Missing values are easy to handle and the variables can be
monitored by plotting the posterior distributions during the learning.  BayesPy
has also preliminary support for some advanced VB methods such as stochastic
variational inference \citep{Hoffman:2013}, deterministic annealing
\citep{Katahira:2008}, collapsed inference \citep{Hensman:2012},
Riemannian conjugate gradient learning \citep{Honkela:2010}, parameter
expansions \citep{Qi:2007} and pattern searches \citep{Honkela:2003}.  For
developers, the unit testing framework helps in finding bugs, making changes and
implementing new features in a robust manner.

@@

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available as
  supplementary material.}  Both models were run for small and large
artificially generated datasets.  For GMM, the small model used 10 clusters for
200 observations with 2 dimensions, and the large model used 40 clusters for
2000 observations with 10 dimensions.  For PCA, the small model used
10-dimensional latent space for 500 observations with 20 dimensions, and the
large model used 40-dimensional latent space for 2000 observations with 100
dimensions.  Infer.NET used a fully factorizing approximation for the PCA model,
whereas BayesPy did not factorize with respect to the latent space dimensions.
The experiments were run on a Linux computer with a quad-core processor
(i7-4702MQ).  Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compare the average CPU time per
iteration.

>>>>

\usepackage{subcaption}
\usepackage{url}
\usepackage{color}
\usepackage{listings}
\usepackage{amsmath}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}

@@

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct a variety of
models, including principal component analysis models, linear state-space
models, mixture models and hidden Markov models.  In addition, BayesPy has been
used in two publications about parameter expansion and time-varying dynamics for
linear state-space models \citep{Luttinen:2013,Luttinen:2014}.
% , which show that it is applicable to large complex models requiring efficient
% inference.

@@

Nodes are the primary building blocks for models in BayesPy.  There are two
types of nodes: stochastic and deterministic.  Stochastic nodes correspond to
probability distributions and deterministic nodes correspond to functions.
Built-in stochastic nodes include all common exponential family distributions
(e.g., Gaussian, gamma and Dirichlet distributions), a general mixture
distribution and a few complex nodes for dynamic variables (e.g., discrete and
Gaussian Markov chains).  Built-in deterministic nodes include a gating node and
a general sum-product node.
%The
%documentation also provides instructions for implementing new nodes in case the
%built-in collection is insufficient.
In case a model cannot be constructed using the built-in nodes, the
documentation provides instructions for implementing new nodes.

BayesPy is designed to be simple enough for non-expert users but flexible and
efficient enough for expert users.  One goal is to keep the syntax easy and
intuitive to read and write by making it similar to the mathematical formulation
of the model.  Missing values are easy to handle and the variables can be
monitored by plotting the posterior distributions during the learning.  BayesPy
has also preliminary support for some advanced VB methods such as stochastic
variational inference \citep{Hoffman:2013}, deterministic annealing
\citep{Katahira:2008}, collapsed inference \citep{Hensman:2012}, Riemannian
conjugate gradient learning \citep{Honkela:2010}, parameter expansions
\citep{Qi:2007} and pattern searches \citep{Honkela:2003}.  For developers, the
unit testing framework helps in finding bugs, making changes and implementing
new features in a robust manner.

@@

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available as
  supplementary material.}  Both models were run for small and large
artificially generated datasets.  For GMM, the small model used 10 clusters for
200 observations with 2 dimensions, and the large model used 40 clusters for
2000 observations with 10 dimensions.  For PCA, the small model used
10-dimensional latent space for 500 observations with 20 dimensions, and the
large model used 40-dimensional latent space for 2000 observations with 100
dimensions.  Infer.NET used a fully factorizing approximation for the PCA model,
whereas BayesPy did not factorize with respect to the latent space dimensions.
The experiments were run on a Linux computer with a quad-core processor
(i7-4702MQ).  Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compared the average CPU time per
iteration.
